\section{Experiments}
\label{sec:experiments}


Our tool generates Alloy code which is then run through the Alloy
Analyzer to do the analysis.  It is targeted for proving things about
$\sqsubseteq$ relations, but we can use it for simpler questions as
well.

One such question might be `give a tuple $e$ such that $\eff (e, p) =
\Permit$' (where $p$ is defined as in our running example).  We
generate the Alloy code for the XACML policy, as normal, and then
append the following:
\begin{verbatim}
fun CheckTuple {
    some T0.permit
}
run CheckTuple for 2 but 2 Bool, 1 Triple, 8 Type
\end{verbatim}
The numbers after \texttt{CheckTuple} establish how deeply we will
look; we're looking for such a tuple in a universe where there are two
Boolean values, only one Triple (\texttt{T0}, the one generated
through $\translation(p)$), 8 domain types (strings, integers, and the
like)), and two elements of everything else.  If we run this, the
analyzer tells us that the tuple $\langle \{ \mtt{Type\_6} \},$ $\{
\mtt{Bool\_1} \},$ $\{ \mtt{Type\_7} \} \rangle$ generates a
\Permit{}, and further examination of the output shows that
\texttt{Bool\_1} is \texttt{True}, \texttt{Type\_6} is 18, and
\texttt{Type\_7} is the string \texttt{vote}.  Through similar means
we can discover that the tuple $\langle \{ 18 \}, \{ \}, \{ \mtt{vote}
\} \rangle$ will generate an error.

For something more interesting, we try to show that $p_v \sqsubseteq_D
p_c$, as we proved manually.  Our coda now becomes
\begin{verbatim}
assert Subset {
    T0.deny in T1.deny
}
check Subset for 2 but 2 Bool, 2 Triple, 10 Type
\end{verbatim}
We get a counterexample almost immediately, giving us the tuple we
constructed in Section~\ref{sec:properties-policies}.  If we modify
the policy so as to restrict result checking to only those who voted
successfully, then the subset relation holds, and no counterexample is
produced.

\subsection{Timing Data}

These are but small examples, so they do not reflect the time required
to solve large problems that one might reasonably ask.  Since the
underlying problem (SAT) is NP complete, it is reasonable to ask
whether these techniques are useful at all as the problem size
increases.

There are two sides to this.  One side is that larger and more complex
policies will inescapably take longer than small policies.  The other
is that---in the context of problems in $BP_3$ and $BP_4$, where we
must restrain the domains to a certain finite size---the amount of
computation involved as the size of the domain increases may trigger
the exponential worst case behavior.

\begin{table*}[t]
\centering
\begin{tabular}{c|cccc|cc|c}
  Property & IO & Transform & CNF & SAT & Variables
  & Clauses & Result \\
  \hline
% unsat
  C1 & 1.85 & 0.17 & 1.35 & 0.11 & 56 & 114 & No \\
  C2 & 2.07 & 0.19 & 1.41 & 0.39 & 42 & 83  & No \\
  C3 & 1.88 & 0.16 & 1.36 & 0.12 & 51 & 108 & No \\
  C4 & 1.94 & 0.17 & 1.33 & 0.11 & 52 & 106 & No \\
  C5 & 1.82 & 0.16 & 2.00 & 0.16 &  79 & 166 & No \\
  \hline
% sat
  C6 & 2.12 & 0.15 & 2.53 & 0.15 & 89 & 190 & Yes \\
  C7 & 2.56 & 0.34 & 3.70 & 0.10 & 95 & 218 & Yes  \\
  C8 & 1.99 & 0.18 & 1.21 & 0.11 & 42 & 83 & Yes \\
  C9 & 1.92 & 0.16 & 1.49 & 0.11 & 51 & 106 & Yes \\
  C10 & 1.88 & 0.19 & 3.47 & 0.11 & 108 & 250 & Yes \\
  C11 & 1.89 & 0.16 & 5.18 & 0.15 & 129 & 297& Yes \\
  \hline
  M1 & 0.75 & 0.02 & 15.10 & 0.22 & 109 & 280 & No \\
  M2 & 1.00 & 0.03 & 14.78 & 0.13 & 108 & 279 & No \\
\hline
  V1 & 0.73 & 0.14 & 5.86 & 0.12 & 52 & 123 & Yes \\
\end{tabular}
\caption{Verification performance for the 
properties of conference manager (C1-11), medico (M1-2) and
voting (V1) examples. 
We divided the execution time to I/O, tranformation
to triple form, CNF transformation and SAT solving.
The times are in seconds.
We also listed the size of the generated SAT problem instance 
(in terms of the number of Boolean variables and clauses)
for verification of each property.
If the result is ``Yes'' the generated SAT instance is satisfiable
and the property is violated. 
If the result is ``No'' the generated SAT instance is not satisfiable
and the property holds.
}
\label{tbl:bigdata}
\end{table*}



To demonstrate the effectiveness of the verification approach we propose 
we have collected 
two sets of data.  The first, which we have charted in
Figure~\ref{fig:graph}, shows the time required to verify or refute a
relationship between two example policies.  It shows how the technique
scales as the size of the domain sets increases.  The second set of
data proves and refutes a relationship between the Medico policy from
Section~4.2 of the XACML specification~\cite{xacml} and a subset of
itself.  This shows that the technique is feasible for larger
policies.  All these benchmarks were performed on a 1 GHz PowerPC, and
all times are the median of five runs, to smooth out irregularities.
In each case, trials were run until the formula proved too large for
Alloy Analyzer to handle; past the given sizes the analyzer would fail
cryptically.  ``Domain size'' means the number of elements we are
using for our analysis, in each domain; so a domain size of 8 for our
example environment $E$ means we simulate every $e \in E$ where
$|e[i]| \leq 8$ for each component in $E$.

The data indicates that the time required for analysis is exponential
in the size of the scope, which is to be expected for a SAT based
algorithm.  However, all times are under two minutes, and our
technique can clearly prove important properties of these problems in
a useful amount of time.

% arch-tag: 5a3a4d97-641f-4c94-804d-9ec83256e4e6

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "paper"
%%% End: 
