\documentclass{acm_proc_article-sp}
\usepackage{amsmath}
\usepackage{times}
\usepackage{mathptm}
\renewcommand{\ttdefault}{cmtt}
\DeclareMathOperator{\eff}{eff}
\DeclareMathOperator{\xpathmatch}{xpathnodematch}
\DeclareMathOperator{\translation}{translate}
\DeclareMathOperator{\extract}{extract}
\DeclareMathOperator{\constant}{constant}
\DeclareMathOperator{\env}{env}
\newcommand{\step}[1]{\intertext{\{#1\}}}
\newcommand{\mtt}[1]{\text{\texttt{#1}}}
\newcommand{\itmathcommand}[1]{\ensuremath{\text{\textit{#1}}}}
\newcommand{\Scope}{\itmathcommand{Scope}}
\newcommand{\Err}{\itmathcommand{Err}}
\newcommand{\Permit}{\itmathcommand{Permit}}
\newcommand{\Deny}{\itmathcommand{Deny}}
\newcommand{\Indeterminate}{\itmathcommand{Indet}}
\newcommand{\NotApplicable}{\itmathcommand{NotApp}}

\begin{document}
\title{Automated Verification of Access Control Policies}
\numberofauthors{2}
\author{
\alignauthor Graham Hughes \quad Tevfik Bultan \\
      \affaddr{Computer Science Department} \\
      \affaddr{University of California} \\
      \affaddr{Santa Barbara, CA 93106, USA} \\
      \email{\{graham,bultan\}@cs.ucsb.edu} \\
}
\date\today
\maketitle

\begin{abstract}
  Managing access control policies in modern computer systems can be
  challenging and error-prone, especially when multiple access
  policies are combined to form new policies, possibly introducing
  unintended consequences.  In this paper we present a framework for
  automated verification of access control policies.  We introduce a
  formal model for systematically specifying access to resources.  We
  show that the access control policies in the XACML access control
  language can be translated to a simple form which partitions the
  input domain to four classes: permit, deny, error, and
  not-applicable.  We present several ordering relations for access
  control policies which can be used to specify the properties of
  the policies and the relationships among them.  We then show how to
  automatically check these ordering relations using an existing
  automated analysis tool.  In particular, we translate XACML policies
  to the Alloy language and check their properties using the Alloy
  Analyzer.  Our experimental results demonstrate that automated
  verification of XACML policies is feasible.
\end{abstract}


\input{introduction}
\input{policyspec}
\input{properties}
\input{translating}
\input{experiments}
\input{conclusion}

\bibliographystyle{abbrv}
%\bibliography{citations/citations.bib}
\begin{thebibliography}{10}

\bibitem{abadpeiro99plas}
J.~L. Abad-Peiro, H.~Debar, T.~Schweinberger, and P.~Trommler.
\newblock {PLAS} --- {P}olicy language for authorizations.
\newblock Technical Report RZ 3126, IBM Research Division, 1999.

\bibitem{abadi93calculus}
M.~Abadi, M.~Burrows, B.~Lampson, and G.~Plotkin.
\newblock A calculus for access control in distributed systems.
\newblock {\em ACM Transactions on Programming Languages and Systems},
  15(4):706--734, September 1993.

\bibitem{bertino98access}
E.~Bertino, C.~Bettini, E.~Ferrari, and P.~Samarati.
\newblock An access control model supporting periodicity constraints and
  temporal reasoning.
\newblock {\em ACM Transactions on Database Systems}, 23(3):231--285, 1998.

\bibitem{bonatti01dataarchives}
P.~Bonatti, E.~Damiani, S.~{De Capitani di Vimercati}, and P.~Samarati.
\newblock An access control model for data archives.
\newblock In {\em Proceedings of the 16th international conference on
  information security: trusted information}, pages 261--276, Paris, France,
  2001. Kluwer International Federation For Information Processing Series.

\bibitem{bonatti02algebra}
P.~Bonatti, S.~{De Capitani di Vimercati}, and P.~Samarati.
\newblock An algebra for composing access control policies.
\newblock {\em ACM Transactions on Information Systems Security}, 5(1):1--35,
  2002.

\bibitem{damiani02SVG}
E.~Damiani, S.~{De Capitani di Vimercati}, E.~Fern\'andez-Medina, and
  P.~Samarati.
\newblock Access control of {SVG} documents.
\newblock In {\em Proceedings of DBSec 2002}, pages 219--230, 2002.

\bibitem{damiani00design}
E.~Damiani, S.~{De Capitani di Vimercati}, S.~Paraboschi, and P.~Samarati.
\newblock Design and implementation of an access control processor for {XML}
  documents.
\newblock {\em Computer Networks}, 33(1--6):59--75, 2000.

\bibitem{damiani02fine}
E.~Damiani, S.~{De Capitani di Vimercati}, S.~Paraboschi, and P.~Samarati.
\newblock A fine-grained access control system for {XML} documents.
\newblock {\em ACM Transactions on Information Systems Security},
  5(2):169--202, 2002.

\bibitem{damiani01controllingaccess}
E.~Damiani, P.~Samarati, S.~{De Capitani di Vimercati}, and S.~Paraboschi.
\newblock Controlling access to {XML} documents.
\newblock {\em IEEE Internet Computing}, 5(6):18--28, 2001.

\bibitem{divimercati96federated}
S.~{De Capitani di Vimercati} and P.~Samarati.
\newblock Access control in federated systems.
\newblock In {\em Proceedings of the 1996 workshop on new security paradigms},
  pages 87--99, Lake Arrowhead, California, United States, 1996. ACM Press.

\bibitem{heckman98applying}
M.~Heckman and K.~N. Levitt.
\newblock Applying the composition principle to verify a hierarchy of security
  servers.
\newblock In {\em {HICSS} (3)}, pages 338--347, 1998.

\bibitem{jackson:micromodels}
D.~Jackson.
\newblock Micromodels of software: Modeling and analysis with {A}lloy.
\newblock \texttt{http://sdg.lcs.mit.edu/alloy/reference-manual.pdf}.

\bibitem{jackson00automating}
D.~Jackson.
\newblock Automating first-order relational logic.
\newblock In {\em Proc. ACM SIGSOFT COnf. Foundations of Software Engineering},
  November 2000.

\bibitem{jackson:elements}
D.~Jackson and C.~A. Damon.
\newblock Elements of style: Analyzing a software design feature with a
  counterexample detector.
\newblock {\em IEEE Transactions on Software Engineering}, 22(7), July 1996.

\bibitem{jackson:alcoa}
D.~Jackson, I.~Schechter, and I.~Shlyakhter.
\newblock Alcoa: the {A}lloy constraint analyzer.
\newblock In {\em Proceedings of International Conference on Software
  Engineering}, Limerick, Ireland, June 2000. IEEE.

\bibitem{jackson01lightweight}
D.~Jackson and A.~Waingold.
\newblock Lightweight extraction of object models from bytecode.
\newblock In {\em IEEE Transactions on Software Engineering}, February 2001.

\bibitem{jajodia01multiplepolicies}
S.~Jajodia, P.~Samarati, M.~L. Sapino, and V.~S. Subrahmanian.
\newblock Flexible support for multiple access control policies.
\newblock {\em ACM Transactions on Database Systems}, 26(2):214--260, 2001.

\bibitem{jajodialogical}
S.~Jajodia, P.~Samarati, and V.~S. Subrahmanian.
\newblock A logical language for expressing authorizations.
\newblock In {\em Proceedings of the 1997 IEEE Symposium on Security and
  Privacy}, pages 31--42, Oakland, CA, USA, 1997. IEEE Press.

\bibitem{jajoida97unified}
S.~Jajodia, P.~Samarati, V.~S. Subrahmanian, and E.~Bertino.
\newblock A unified framework for enforcing multiple access control policies.
\newblock In {\em SIGMOD'97}, pages 474--485, Tucson, AZ, May 1997.

\bibitem{marinov:exhaustive}
D.~Marinov, A.~Andoni, D.~Danilinc, S.~Khurshid, and M.~Rinard.
\newblock An evaluation of exhaustive testing for data structures.
\newblock Technical Report MIT-LCS-TR-921, MIT CSAIL, 2003.

\bibitem{marinov02valloy}
D.~Marinov and S.~Khurshid.
\newblock {VA}lloy: Virtual functions meet a relational language.
\newblock In {\em 11th International Symposium of Formal Methods Europe (FME
  2002)}, Copenhagen, Denmark, July 2002.

\bibitem{naumovich:permission-analysis}
G.~Naumovich.
\newblock A conservative algorithm for computing the flow of permissions in
  {J}ava programs.
\newblock In {\em Proceedings of the International Symposium on Software
  Testing and Analysis ({ISSTA} '02)}, pages 33--43, July 2002.

\bibitem{nolte02filesystem}
T.~Nolte.
\newblock Exploring filesystem synchronization with lightweight modeling and
  analysis.
\newblock Master's thesis, MIT, August 2002.

\bibitem{samarati01policiesmodelsmechanisms}
P.~Samarati and S.~{De Capitani di Vimercati}.
\newblock {\em Foundations of Security Analysis and Design}, chapter~3, pages
  137--196.
\newblock Springer Verlag, 2001.

\bibitem{sandhu96audit}
R.~Sandhu and P.~Samarati.
\newblock Authentication, access control, and audit.
\newblock {\em ACM Computing Surveys}, 28(1):241--243, 1996.

\bibitem{sandhu93latticebased}
R.~S. Sandhu.
\newblock Lattice-based access control models.
\newblock {\em IEEE Computer}, 26(11):9--19, November 1993.

\balancecolumns
\bibitem{sandhu96rolebased}
R.~S. Sandhu, E.~J. Coyne, H.~L. Feinstein, and C.~E. Youman.
\newblock Role-based access control models.
\newblock {\em IEEE Computer}, 29(2):38--47, 1996.

\bibitem{sandhu94access}
R.~S. Sandhu and P.~Samarati.
\newblock Access control: Principles and practice.
\newblock {\em IEEE Communications Magazine}, 32(9):40--48, 1994 1994.

\bibitem{schaad:lightweight}
A.~Schaad and J.~Moffet.
\newblock A lightweight approach to specification and analysis of role-based access control extensions.
\newblock In {\em 7th ACM Symposium on Access Control Models and Technologies (SACMAT 2002)}, June 2002.

\bibitem{vaziri03heap}
M.~Vaziri and D.~Jackson.
\newblock Checking heap-manipulating procedures with a constraint solver.
\newblock In {\em TACAS'03}, Warsaw, Poland, 2003.

\bibitem{waingold01automated}
A.~Waingold.
\newblock Automated extraction of abstract object models.
\newblock Master's thesis, MIT, May 2001.

\bibitem{zao:rbac}
J.~Zao, H.~Wee, J.~Chu, and D.~Jackson.
\newblock {RBAC} schema verification using lightweight formal model and constraint analysis.
\newblock XXX

\bibitem{xacml}
{eXtensible Access Control Markup Language (XACML)} version 1.0.
\newblock OASIS Standard, February 2003.

\bibitem{XML}
Extensible markup language ({XML}) 1.0 (second edition), 2000.

\bibitem{xmlschema}
{XML Schema} part 2: Datatypes.
\newblock W3C Recommendation, May 2001.

\end{thebibliography}

\clearpage
\appendix
\section{XACML Representation of Voting Example and Alloy Translation}
\label{sec:xacml-repr-voting}

{ \small
\input{trimmedexample.xacml.tex}
}
\newpage

\section{Alloy Translation}
\label{sec:alloy-trans}

{ \small
\input{trimmedexample.als.tex}
}

\end{document}
